{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-11-23 thesis/landslide.tex
dave revisions on landslide chapater

The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of Chapter~\ref{chap:quicksand} to itself.
% TODO: add quicksand symlink
This is done via the {\tt ./landslide} script in the top-level directory, which:
\begin{itemize}
	\item Checks if the user needs to run {\tt *-setup.sh} again, in case her source code was more recently updated than the existing annotated build (a common mistake),
	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
		and
	\item (If during the student user study,) compresses the resulting log files,
		creates a snapshot tarball of them and the current version of the user's code,
		and sends it to me for nefarious research purposes.
\end{itemize}

@@

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C}).
		According to \sect{\ref{sec:quicksand-soundness}}'s soundness proofs,
		this is equivalent to {\tt -0}
		(and according to my experience, way faster as well).
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

Use of {\tt tell\_landslide()} annotations is not necessary,
although {\tt tell\_landslide\_\allowbreak{}preempt()} and {\tt tell\_landslide\_dump\_stack()}
may optionally be used at the user's convenience.
Additionally, the following ``secret'' annotations
are occasionally used in the pre-supplied test cases
to accomplish several mysterious goals described hereupon.

@@

Test cases may define global variables of the following names
to instruct Landslide to assert the following corresponding predicates at the end of each test execution,
after all threads exit.
Each predicate will be checked iff its first listed variable name is defined;
if that variable is defined, all others associated must also be;
any combination of the three first-listed variables may be specified at the user's option.
\begin{itemize}
	\item {\tt magic\_global\_expected\_result == magic\_global\_value}
	\item {\tt magic\_expected\_sum == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child}
	\item {\tt magic\_expected\_sum\_minus\_result == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child - magic\_global\_value}
\end{itemize}
These could not be implemented as asserts in the test code itself
without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()},
avoiding which is important for tests to be student-accessible earlier in the project implementation timeline.

@@

When Landslide finds a bug, it produces an execution trace of the particular interleaving of threads that led to the bug.
This takes the form of a two-dimensional table,
with a column for each thread,
and each row representing the continuous execution of one thread between two (not necessarily consecutive) preemption points.
In each row, the cell in the column corresponding to the executed thread will contain a stack trace,
indicating the code location of the preemption point {\em at the end} of that thread transition
(i.e., each stack trace indicates ``this thread ran until it reached the indicated line of code'').
The bug reports are formatted in html, recommended to be viewed in a web browser.
An example is shown in Figure~\ref{fig:bugreport}.

In addition to the preemption trace, the bug report provides some additional helpful information:
a stack trace of the current thread at the ultimate point when the bug was executed,
a message indicating the nature of the bug encountered,
statistics about the size of the state space,
and optionally additional information about the bug.%
\footnote{For certain types of bugs, not pictured here;
for example, use-after-frees will report separate stack traces
indicating when the corresponding heap block was last allocated and freed.
The intrepid source-diver may find all such cases of extra bug details
by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}

@@

The educational experiments in this thesis focus on projects which students implement on top of provided kernel basecode which Landslide already ``understands''.
Such understanding is conferred via the annotations described in this section.
For P2 and Pintos students I supply these annotations behind the scenes,
but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.

@@

@@

		Some kernels\footnote{most, actually} remove the current thread from their runqueue,

@@

		Other kernels\footnote{the author's own student kernel from long ago}

@@

@@

		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \${}6.2.3.}

@@

		\footnote{once used in the bad old days; now recommended for debugging use only}

@@

The following annotations enable Landslide to track locksets for data race analysis.
If not provided, it will be as if Landslide assumes no guarantees about mutual exclusion or happens-before,
and hence will identify all memory conflicts as data races.
(Note that the corresponding instrumentation for P2s is achieved automatically,
as the names of the mutex interface are mandated by the project specification.)

@@

The Landslide scheduler, which lives in {\tt schedule.c}, has two main duties:
to maintain an accurate representation of all the existing threads on the simulated system
and track what concurrency-relevant actions each is performing at any given time,
and to orchestrate the sequence of timer interrupts necessary
to cause the simulated system to context switch to any given thread at any given time.
System-wide state is stored in a single {\tt struct sched\_state},
including the thread queues (runqueue, deschedule queue, and sleep queue),
while per-thread state is stored in {\tt struct agent}s (named after the terminology of \cite{dbug-ssv}) which live on said queues.

@@

\begin{itemize}
	\item {\tt cause\_timer\_interrupt()}
		triggers a pending timer interrupt, whose handler will be entered as soon as the upcoming instruction is executed.
	\item {\tt cause\_timer\_interrupt\_immediately()}
		does as above, but forces the system to enter the interrupt handler before the upcoming instruction is executed.
		That instruction will be executed upon return from the interrupt.
	\item {\tt avoid\_timer\_interrupt\_immediately()}
		suppresses a timer interrupt triggered by the simulator from outside of Landslide's control.
		It acknowledges the APIC and forces the system to jump to the end of the interrupt handler.
	\item {\tt delay\_instruction()}
		forces the system to execute a no-op before the upcoming instruction,
		effectively converting an invocation of {\tt cause\_timer\_interrupt()} to {\tt cause\_timer\_interrupt\_immediately()}.
	\item {\tt cause\_keypress()} triggers a keyboard event corresponding to the specified character.
		The interrupt will be taken after the upcoming instruction is executed
		(provided no timer interrupt is simultaneously pending).
		Only {\tt a-z}, {\tt 0-9}, {\tt \_}, space, and newline are supported (enough to name any Pebbles test case).
	\item {\tt interrupts\_enabled()} queries the CPU's interrupt flag ({\tt eflags:IF}).
	\item {\tt cause\_transaction\_failure()} forces {\tt \_xbegin()} to return a specified abort code.
\end{itemize}

@@

After the explorer picks a past point of the program to preempt, % sorry this started out like that and i just had to roll with it
Landslide collaborates with the simulator to revert the machine state to that point before switching to the desired thread.
The Simics version is merely a bunch of wrapper glue code around the {\tt set-bookmark} and {\tt skip-to} backtracking commands.
Bochs however does not support backtracking, so I instead use {\tt fork()} to get Linux to copy the machine state for me.

@@

The options specified in \sect{\ref{sec:landslide-directly}}
are handled by a family of gross shell scripts that live in {\tt pebsim/}.

@@

In this section $t_i$ will denote a transition between two program states during execution,
with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
and $T(t_i)$ will denote the thread which was scheduled to produce that transition.
A visual example will be given at the end to help reinforce the intuition behind the formalism.

@@

The happens-before relation expresses when two thread transitions can potentially be reordered,
or in other words, are concurrent (despite the serialized nature of the simulated execution).
This relation is expressed in the following definitions paraphrased from \cite{dpor}.

\begin{itemize}
	\item {\bf Enabled}: A transition $t_i$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t_i} s'$ exists.
		In systems terms, the scheduler at state $s$ considers $T(t_i)$ to be runnable.
	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
		\begin{enumerate}
			\item $t_1$ is enabled in $s$ and $s \xrightarrow{t_1} s'$, and
			\item $t_2$ is enabled in $s$ but not enabled in $s'$, or vice versa.
		\end{enumerate}
		In systems terms, either $T(t_1) = T(t_2)$,
		or the execution of $t_1$ at $s$ causes $T(t_2)$
		to change state from blocked to runnable or vice versa.
		\footnote{The original paper's definition includes a second criterion that, from $s$,
		%the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
		a state $s'$ exists such that both $s \xrightarrow{t_1 t_2} s'$ and $s \xrightarrow{t_2 t_1} s'$.
		This captures the memory independence relation,
		but computationally requires direct comparison of program states.
		{\em Stateless} model checkers compute memory conflicts separately from happens-before,
		to find and prune such identical states implicitly,
		as described in the next two subsections.}
		Landslide computes this relation in {\tt enabled\_by()} in {\tt save.c}.
	\item {\bf Happens-Before}: The happens-before relation for a transition sequence $S = t_1 \dots t_n$
		is the smallest relation $\rightarrow_S$ on ${1 \dots n}$ such that
		\begin{enumerate}
			\item if $i < j$ and $S_i$ and $S_j$ are dependent then $i \rightarrow_S j$, and
			\item $\rightarrow_S$ is transitively closed.
		\end{enumerate}
		Landslide computes this relation in
		%the aptly-named
		{\tt compute\_happens\_before()} in {\tt save.c}.
\end{itemize}
The happens-before relation is a partial order expressing the scheduling constraints of a given interleaving.
All pairs of interleavings not included are subject to reordering,
and hence candidates for new interleavings to test.

@@

Whenever a conflict is identified, it also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
It checks for {\tt free()}/access conflicts as well as access pairs,
effectively treating deallocation of a heap block as a ``poisoning'' write to its entire contents,
which is considered to conflict with accesses to any address therein
on the grounds that reordering may expose a use-after-free.

@@

{\bf Identifying new interleavings.}
To find which alternate interleavings need to be tested,
% TODO: change i<4 to i<j.
% i accidentally typed i<4 here and decided it'd be funny to leave it this way
DPOR compares pairwise each pair of transitions $t_i$ and $t_j$, $i<4$, in the current interleaving $S$.
If $t_i \rightarrow_S t_j$ then they cannot be reordered,
and if $t_i \not\leftrightsquigarrow_S t_j$ then reordering them
will produce a state already encountered in this interleaving;
hence, DPOR marks new interleavings only
when $t_i \not\rightarrow_S t_j$ and $t_i \leftrightsquigarrow_S t_j$
%for transition pairs not related by happens-before and are related by memory conflict
({\tt is\_evil\_ancestor()}).
\footnote{To aid intuition, consider the two extremes:
if all transitions are related by happens-before,
the program is not concurrent and no alternate interleavings are possible;
if all transitions are memory-independent,
the program exhibits full data isolation between threads and all schedules are observably equivalent.}

For each such pair, let $s$ denote the state (preemption point) before $t_i$.
Then:
\begin{itemize}
	\item If $T(t_j)$ is runnable at $s$, return $I_{ij} = (t_1 \dots t_{i-1}, T_j)$ ({\tt tag\_good\_sibling()}).
	\item Otherwise, there must be some third thread runnable at $s$;
		\footnote{AFSOC $T(t_i)$ is the only runnable thread at $s$,
		then either $T(t_i)$'s execution at $s$ enabled $T(t_j)$,
		or it enabled an intermediate transition
		(whether by $T(t_i)$ or a third thread)
		which in turn enabled $T(t_j)$.
		In either case $t_i$ is transitively dependent with $t_j$, contradicting $t_i \not\rightarrow_S t_j$.}
		%(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that
		$T_k \ne T(t_i)$ and $T_k$ is runnable at $s$
		({\tt tag\_all\_siblings()}).
\end{itemize}

@@

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=\textwidth]{dpor-example-0.pdf}
	\end{center}
	\caption{Result of a single iteration of DPOR.}
	\label{fig:dpor-example-0}
\end{figure}

@@

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered \dporTB{{\tt tmp2 <- x}} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting \dporTB{{\tt x <- tmp2}}.

@@

To see why branches D through J need not be tested, consider that each thread's {\tt tmpN++} is a thread-local event,
participating in no memory conflicts,
and hence any two interleavings differing only by reordering {\tt tmpN++}s must be equivalent.
The dashed blue arrows denote such equivalences;
note the two disjoint equivalence classes \{B,E,G,I\} and \{C,D,F,H,J\},
distinguished by the order of the two final {\tt x <- tmpN}s.
Note also that
%(as shown in Figure~\ref{fig:tree}(b)),
although B and C also have the same outcome ({\tt x==1}),
this depends on the {\em values} written to memory rather than {\em addresses}
(and would change if one thread's {\tt tmpN++} were a {\tt tmpN+=2}, for example),
which DPOR does not consider.
Recent work \cite{mcr} has extended DPOR to find such value-based equivalences,
although is beyond this explanation's scope.

@@

In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
As the presence of more equivalence arrows in Figure~\ref{fig:dpor-example-2} hints,
its reduced subset state space still contains redundancy,
arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
Visual inspection shows that $\bigstar$4, while locally justified
in trying to reorder \dporTA{{\tt tmp1 <- x}} before \dporTB{{\tt x <- tmp2}},
effectively serves % oops
only to reorder it with \dporTB{{\tt tmp2 <- x}}
relative to the first symmetric subtree ($\bigstar$1).
In other words,
even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
Figure~\ref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.

@@

In Pure Happens-Before,
condition 4 is replaced with the traditional distributed systems notion of Happens-Before \cite{lamport-clocks}.
Landslide implements this via the vector clocks approach described by {\sc FastTrack} \cite{fasttrack}.
I refer the reader interested in the vector clock algorithm itself to the {\sc FastTrack} paper,
limiting discussion here to Landslide's corresponding implementation of each inference rule.

>>>>

The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of Chapter~\ref{chap:quicksand} to itself.
% TODO: add quicksand symlink
This is done via the {\tt ./landslide} script in the top-level directory, which:
\begin{itemize}
	\item Checks if the user needs to run {\tt *-setup.sh} again, in case her source code was more recently updated than the existing annotated build (a common mistake),
	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
		and
	\item (If during the student user study,) compresses the resulting log files,
		creates a snapshot tarball of them and the current version of the user's code,
		and sends it to me for nefarious research purposes
		\revisionminor{(see \sect{\ref{sec:education-pebbles-instrumentation}})}.
\end{itemize}

@@

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C}).
		According to \sect{\ref{sec:quicksand-soundness}}'s soundness proofs,
		this is equivalent to {\tt -0}
		(and according to my experience, \revisionminor{much} faster as well).
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

Use of {\tt tell\_landslide()} annotations
\revisionminor{(\sect{\ref{sec:tell-landslide}})}
is not necessary,
although {\tt tell\_landslide\_\allowbreak{}preempt()} and {\tt tell\_landslide\_dump\_stack()}
may optionally be used at the user's convenience.
Additionally, the following ``secret'' annotations
are occasionally used in the pre-supplied test cases
to accomplish several mysterious goals described hereupon.

@@

Test cases may define global variables of the following names
to instruct Landslide to assert the following corresponding predicates at the end of each test execution,
after all threads exit.
Each predicate will be checked iff its first listed variable name is defined;
if that variable is defined, all others associated must also be;
any combination of the three first-listed variables may be specified at the user's option.
\begin{itemize}
	\item {\tt magic\_global\_expected\_result == magic\_global\_value}
	\item {\tt magic\_expected\_sum == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child}
	\item {\tt magic\_expected\_sum\_minus\_result == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child - magic\_global\_value}
\end{itemize}
\revision{Because Landslide tests the ultimate value of these variables after all threads have completed execution,}
these could not be implemented as asserts in the test code itself
without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()},
avoiding which is important for tests to be student-accessible earlier in the project implementation timeline.

@@

When Landslide finds a bug, it produces an execution trace of the particular interleaving of threads that led to the bug.
This takes the form of a two-dimensional table,
with a column for each thread,
and each row representing the continuous execution of one thread between two (not necessarily consecutive) preemption points.
In each row, the cell in the column corresponding to the executed thread will contain a stack trace,
indicating the code location of the preemption point {\em at the end} of that thread transition
(i.e., each stack trace indicates ``this thread ran until it reached the indicated line of code'').
The bug reports are formatted in \revisionminor{HTML}, recommended to be viewed in a web browser.
An example is shown in Figure~\ref{fig:bugreport}.

In addition to the preemption trace, the bug report provides some additional helpful information:
a stack trace of the current thread at the ultimate point when the bug was executed
\revision{(the same as the stack trace in the cell corresponding to that thread in the bottom row of the table)},
a message indicating the nature of the bug encountered,
statistics about the size of the state space,
and optionally additional information about the bug.%
\footnote{For certain types of bugs, not pictured here;
for example, use-after-frees will report separate stack traces
indicating when the corresponding heap block was last allocated and freed.
The intrepid source-diver may find all such cases of extra bug details
by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
\revision{\sect{\ref{sec:future-friendly}} discusses how these bug reports might be further improved
by adding more information still,
such as identifying data-race preemption points or listing memory conflicts that occurred during each transition.}

@@

\revision{My M.S. thesis \cite{landslide}
investigated the annotation overhead required for a user to instrument a Pebbles kernel for use with Landslide.
On average, students who volunteered for the study (conducted during P3) required 2 hours for this process.
While many of these students then went on to find and diagnose bugs,
this was deemed an unacceptable burden for an educational tool to impose on struggling students.
%(as opposed to those who had enough free time and ambition to volunteer for a study
%that was advertised in advance as requiring manual annotations)

Hence,}
the educational experiments in this thesis focus on projects which students implement
on top of provided kernel basecode which Landslide already ``understands''.
Such understanding is conferred via the annotations described in this section.
For P2 and Pintos students I supply these annotations behind the scenes,
but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.

@@

\revision{Those defined with an equals sign between name and value
are implemented as bash variables,
which the annotation scripts check after processing the configuration to emit a corresponding {\tt \#define}
within Landslide itself;
those defined with no equals sign
are implemented as bash functions,
which the annotation scripts define in advance of processing the configuration to do something more complicated.}

@@

		Some kernels\footnote{Most, actually.} remove the current thread from their runqueue,

@@

		Other kernels\footnote{The author's own student kernel from long ago.}

@@

		\revision{Whether or not a kernel requires this annotation could be auto-detected in future work.}

@@

		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \sect{6.2.3}.}

@@

		\footnote{Once used in the bad old days; now recommended for debugging use only.}

@@

The following annotations enable Landslide to track locksets~\cite{eraser} for data race analysis.
If not provided, it will be as if Landslide assumes no guarantees about mutual exclusion or happens-before,
and hence will identify all memory conflicts as data races.
(Note that the corresponding instrumentation for P2s is achieved automatically,
as the names of the mutex interface are mandated by the project specification.)

@@

The Landslide scheduler, which lives in {\tt schedule.c}, has two main duties:
to maintain an accurate representation of all the existing threads on the simulated system
and track \revisionminor{which} concurrency-relevant actions each is performing at any given time,
and to orchestrate the sequence of timer interrupts necessary
to cause the simulated system to context switch to any given thread at any given time.
System-wide state is stored in a single {\tt struct sched\_state},
including \revisionminor{several queues to track threads in various states of runnability}
(runnable queue, deschedule queue, and sleep queue),
while per-thread state is stored in {\tt struct agent}s (named after the terminology of \cite{dbug-ssv}) which live on said queues.

@@

\begin{itemize}
	\item {\tt cause\_timer\_interrupt()}
		triggers a pending timer interrupt, whose handler will be entered as soon as
		\revisionminor{the execution of the upcoming instruction is completed}.
	\item {\tt cause\_timer\_interrupt\_immediately()}
		does as above, but forces the system to enter the interrupt handler before the upcoming instruction is executed.
		That instruction will be executed upon return from the interrupt.
	\item {\tt avoid\_timer\_interrupt\_immediately()}
		suppresses a timer interrupt triggered by the simulator from outside of Landslide's control.
		It acknowledges the APIC and forces the system to jump to the end of the interrupt handler.
	\item {\tt delay\_instruction()}
		forces the system to execute a no-op before the upcoming instruction,
		effectively converting an invocation of {\tt cause\_timer\_interrupt()} to {\tt cause\_timer\_interrupt\_immediately()}.
	\item {\tt cause\_keypress()} triggers a keyboard event corresponding to the specified character.
		The interrupt will be taken after the upcoming instruction is executed
		(provided no timer interrupt is simultaneously pending).
		Only {\tt a-z}, {\tt 0-9}, {\tt \_}, space, and newline are supported (enough to name any Pebbles test case).
	\item {\tt interrupts\_enabled()} queries the CPU's interrupt flag ({\tt eflags:IF}).
	\item {\tt cause\_transaction\_failure()} forces {\tt \_xbegin()} to return a specified abort code.
\end{itemize}

@@

After the explorer picks a past point of the program to preempt, % sorry this started out like that and i just had to roll with it
Landslide collaborates with the simulator to revert the machine state to that point before switching to the desired thread.
The Simics version is merely a bunch of wrapper glue code around the {\tt set-bookmark} and {\tt skip-to} backtracking commands.
Bochs however does not support backtracking, so I instead use {\tt fork()} to get Linux
to copy the \revisionminor{Bochs process and thus the simulation} state for me.

@@

The options specified in \sect{\ref{sec:landslide-directly}}
are handled by a family of \revisionminor{distasteful} shell scripts that live in {\tt pebsim/}.

@@

In this section $t_i$ will denote a transition between two program states during execution,
with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
and $T(t_i)$ will denote the thread which was scheduled \revisionminor{(switched to)} to produce that transition.
A visual example will be given at the end to help reinforce the intuition behind the formalism.

@@

The happens-before relation expresses when two thread transitions can potentially be reordered,
or in other words, are \revisionminor{logically} concurrent (despite the serialized nature of the simulated execution).
This relation is expressed in the following definitions paraphrased from \cite{dpor}.

\begin{itemize}
	\item {\bf Enabled}: A transition $t_i$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t_i} s'$ exists.
		In systems \revisionminor{research} terms, the scheduler at state $s$ considers $T(t_i)$ to be runnable.
	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
		\begin{enumerate}
			\item $t_1$ is enabled in $s$ and $s \xrightarrow{t_1} s'$, and
			\item $t_2$ is enabled in $s$ but not enabled in $s'$, or vice versa.
		\end{enumerate}
		In systems terms, either $T(t_1) = T(t_2)$,
		or the execution of $t_1$ at $s$ causes $T(t_2)$
		to change state from blocked to runnable or vice versa.%
		\footnote{The original paper's definition includes a second criterion that, from $s$,
		%the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
		a state $s'$ exists such that both $s \xrightarrow{t_1 t_2} s'$ and $s \xrightarrow{t_2 t_1} s'$.
		This captures the memory independence relation,
		but computationally requires direct comparison of program states.
		{\em Stateless} model checkers compute memory conflicts separately from happens-before,
		to find and prune such identical states implicitly,
		as described in the next two subsections.}
		Landslide computes this relation in {\tt enabled\_by()} in {\tt save.c}.
	\item {\bf Happens-Before}: The happens-before relation for a transition sequence $S = t_1 \dots t_n$
		is the smallest relation $\rightarrow_S$ on ${1 \dots n}$ such that
		\begin{enumerate}
			\item if $i < j$ and $S_i$ and $S_j$ are dependent then $i \rightarrow_S j$, and
			\item $\rightarrow_S$ is transitively closed.
		\end{enumerate}
		Landslide computes this relation in
		%the aptly-named
		{\tt compute\_happens\_before()} in {\tt save.c}.
\end{itemize}
The happens-before relation is a partial order expressing the scheduling constraints of a given interleaving.
All pairs of interleavings not included are subject to reordering,
and hence candidates for new interleavings to test.

@@

Whenever a conflict is identified,
\revisionminor{Landslide} also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
It checks for {\tt free()}/access conflicts as well as access pairs,
effectively treating deallocation of a heap block as a ``poisoning'' write to its entire contents,
which is considered to conflict with accesses to any address therein
on the grounds that reordering may expose a use-after-free.

@@

{\bf Identifying new interleavings.}
To find which alternate interleavings need to be tested,
DPOR compares pairwise each pair of transitions $t_i$ and $t_j$, $i<j$, in the current interleaving $S$.
If $t_i \rightarrow_S t_j$ then they cannot be reordered,
and if $t_i \not\leftrightsquigarrow_S t_j$ then reordering them
will produce a state already encountered in this interleaving;
hence, DPOR marks new interleavings only
when $t_i \not\rightarrow_S t_j$ and $t_i \leftrightsquigarrow_S t_j$
%for transition pairs not related by happens-before and are related by memory conflict
({\tt is\_evil\_ancestor()}).
\footnote{To aid intuition, consider the two extremes:
if all transitions are related by happens-before,
the program is not concurrent and no alternate interleavings are possible;
if all transitions are memory-independent,
the program exhibits full data isolation between threads and all schedules are observably equivalent.}

For each such pair, let $s$ denote the state (preemption point) before $t_i$.
Then:
\begin{itemize}
	\item If $T(t_j)$ is runnable at $s$, return $I_{ij} = (t_1 \dots t_{i-1}, T_j)$ ({\tt tag\_good\_sibling()}).
	\item Otherwise, there must be some third thread runnable at $s$;%
		\footnote{AFSOC $T(t_i)$ is the only runnable thread at $s$,
		then either $T(t_i)$'s execution at $s$ enabled $T(t_j)$,
		or it enabled an intermediate transition
		(whether by $T(t_i)$ or a third thread)
		which in turn enabled $T(t_j)$.
		In either case $t_i$ is transitively dependent with $t_j$, contradicting $t_i \not\rightarrow_S t_j$.}
		%(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that
		$T_k \ne T(t_i)$ and $T_k$ is runnable at $s$
		({\tt tag\_all\_siblings()}).
\end{itemize}

@@

% TODO: check figure placement (should be exactly in line with text) - same for subsequent 2
% TODO: maybe it's ok t ohave them not in line?
\begin{figure}[h]
	\begin{center}
		\includegraphics[width=\textwidth]{dpor-example-0.pdf}
	\end{center}
	\caption{Result of a single iteration of DPOR.}
	\label{fig:dpor-example-0}
\end{figure}

@@

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered \dporTB{{\tt tmp2~<-~x}} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting \dporTB{{\tt x~<-~tmp2}}.

@@

To see why branches D through J need not be tested, consider that each thread's {\tt tmpN++} is a thread-local event,
participating in no memory conflicts,
and hence any two interleavings differing only by reordering {\tt tmpN++}s must be equivalent.
The dashed blue arrows denote such equivalences;
note the two disjoint equivalence classes \{B,E,G,I\} and \{C,D,F,H,J\},
distinguished by the order of the two final {\tt x~<-~tmpN}s.
Note also that
%(as shown in Figure~\ref{fig:tree}(b)),
although B and C also have the same outcome ({\tt x==1}),
this depends on the {\em values} written to memory rather than {\em addresses}
(and would change if one thread's {\tt tmpN++} were a {\tt tmpN+=2}, for example),
which DPOR does not consider.
Recent work \cite{mcr} has extended DPOR to find such value-based equivalences,
although is beyond this explanation's scope.

@@

In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
As the presence of more equivalence arrows in Figure~\ref{fig:dpor-example-2} hints,
its reduced subset state space still contains redundancy,
arising from the fact that one pair of those \revisionminor{four} events is two reads, and hence not actually in conflict.
Visual inspection shows that $\bigstar$4, while locally justified
in trying to reorder \dporTA{{\tt tmp1~<-~x}} before \dporTB{{\tt x~<-~tmp2}},
effectively serves % oops
only to reorder it with \dporTB{{\tt tmp2~<-~x}}
relative to the first symmetric subtree ($\bigstar$1).
In other words,
even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
Figure~\ref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.

@@

In Pure Happens-Before,
condition \#4 is replaced with the traditional distributed systems notion of Happens-Before \cite{lamport-clocks}.
Landslide implements this via the vector clocks approach described by {\sc FastTrack} \cite{fasttrack}.
I refer the reader interested in the vector clock algorithm itself to the {\sc FastTrack} paper,
limiting discussion here to Landslide's corresponding implementation of each inference rule.
